Journey: A Study of Scalabilty of Memory Management in Linux

### 2017-04-29 paper.tex
Address todos

@@

Currently, there are few good tools available to systems researchers for
understanding how systems interact with large memories. We see the need for a
general lightweight technique or tool for studying and designing operating
systems capable of handling large memory while using systems currently
available to researchers.  An important first step towards building such a tool
is to identify bottlenecks in managing large memories that make a tool
difficult to develop.

In particular, we look at the case of a single process using a large amount of
memory in Linux. We run experiments to identify the costs of
\texttt{vm\_area\_struct}, a common bookkeeping data structure in the Linux
kernel, along with the memory and processing overheads of swapping. From these
data, we extrapolate to systems running with more memory with an eye on
implications for studying large systems. While these experiments are not
comprehensive, we believe they can offer insights that are useful both for
building a research tool and for building systems.

@@

A process may use the \texttt{mmap} system call to request that the kernel
allocate part of its virtual address space. Accesses to memory addresses
outside of allocated regions result in a segmentation fault. This makes memory
corruption errors easier to find and debug, and increases security from memory
corruption attacks.

The Linux kernel keeps track of which portions of a process's address space
have been allocated using the \texttt{vm\_area\_struct} data structure. Because
processes may map a number of regions, including shared objects, stack, heap,
text, and anonymous regions, the scalability of this structure is critical to
Linux's performance.

@@

We measure the memory overhead and latency of the Linux Memory Management
system's use of \texttt{vm\_area\_struct}, which represents a region of a
process's virtual address space. All the physical pages in the system are
managed using the \texttt{struct page} descriptors in the kernel.  We do not
measure the overhead of \texttt{struct page} because it is known that there is
one struct for each physical page. We do not measure the overhead of page
tables because this can be calculated from the size of the virtual memory space
used by the process.  When a program makes a memory management system call
such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap}, \texttt{mprotect},
\texttt{mlock}, or \texttt{madvise}, the kernel is creating, changing, or
removing \texttt{vm\_area\_struct}s; thus, measuring the scalability of Linux
memory management requires measuring the overhead of these structs.

@@

Before swapping a page out to disk, the kernel must reserve a slot for the page
in the swap space, unmap it from the process's address space, and record which
swap slot corresponds to the page.

@@

The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
intervals and sleeps in between. Also, it may be invoked if the system detects
that memory has gone below a threshold. kswapd will continue to run until enough
pages are freed to bring the free memory pool above some threashold. Then,
kswapd goes back to sleep.  One concern with large memory systems is that kswapd
may expend significant CPU time scanning through hundreds of thousands of pages.

@@

To measure memory usage, we use the procfs's reporting on current memory usage.
Where appropriate, we record the slab allocator's count of active
\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
associating memory usage with processes is difficult. Our methodology assumes
that our benchmarks are running in an otherwise idle system and should dominate
increases in memory usage.  We compute total memory usage by the rest of the
system as the total amount of memory less the amount of memory used by the
benchmark. We keep transparent huge pages on throughout our experiments, unless
otherwise specified.

@@

First, when a continuous \texttt{mmap} (with the same permissions) happens the
kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
Instead, it uses the existing \texttt{vm\_area\_struct} to account for the
newly allocated space. This obviates the need to use the slab allocator to create
a new struct.

@@

The kernel latency of creating a memory region is lowest when the new memory is
allocated next to an already allocated region. This latency is the worst in the
\texttt{frag} microbenchmark i.e., when an already mapped region is to be expanded. We
suspect a lot of real world applications to show the fragmented behavior since
it is common for applications to exapnd their existing data structures.  This
finding highlights the imporance of optimistically leaving empty spaces around
the allocated regions to speed up memory expension. This practice would
particularly be useful for performance crucial applications running atop compute
infrastructures where memory is not a scarce resource, but kernel is.

(TODO: (markm) I am hesitant about these conclusions...)

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figures/mmap_latency}
    \caption{Latency of memory operations}
    \label{fig:mmap_latency}
\end{figure}

@@

The \texttt{rand} workload, on the other hand, writes to pages that are likely not
adjacent and likely not subpages of a single huge page. As a result, a huge page
is allocated with almost every write, and system memory is exhausted and begins
swapping very quickly. Memory made available by kswapd is quickly consumed by
further allocations. Hence, the memory used by the workload continues to stay
close to its peak value. As expected, the behavior is the same irrespective of
the size of the swap file.

@@

The \texttt{cont} benchmark has more surprising results. Notice that Figure
\ref{fig:swap_time_cont} has a logarithmic y-scale. The system only begins to
swap after about 4M operations, as before. Unlike the \texttt{rand} benchmark, we see a
number of operations that take hundreds of milliseconds. Figure
\ref{fig:swap_time_cont_cdf} shows that these only form a fraction of a percent
of all accesses done after swapping starts. However, Figure
\ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
time; they impact performance significantly. (TODO: why do they happen?).

@@

@@

I would tell you a UDP joke, but you might not get it.

@@

@@

@@

Likewise, the Quartz emulator studies the behaviour of applications on
Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
existing hardware capabilities to emulate the higher latency and lower bandwidth
of most NVM technologies. While Quartz does not address the fact that the actual
memory available to the applications on NVM might be orders of magnitude higher
than on DRAM, it demonstrates that new technologies may be emulated using
existing technology \cite{quartz}. (TODO: is this system still relavent?)

Two threads walk into a bar. The barkeeper looks up and yells, “hey, I want
don’t any conditions race like time last!“

@@

Interestingly, many of non-volatile memory technologies are making their ways
into storage drives, such as Intel's recently-announced Optane SSDs. One
potential use of fast storage devices is for fast swapping space. Some
literature has proposed that small amounts of DRAM backed by large amounts of
fast swap space can be a scalable alternative to having large amounts of fast
RAM (TODO: cite). (TODO: can we have this in our report?). In such a system,
the overhead of swapping may become a bottleneck.

Prefetching pages from swap space can offer a way to mitigate the overhead of
memory overcommitment, possibly making large virtual machines a viable
mechanism for studying large memory systems. When there is significant memory
pressure, even pages which are likely to be accessed soon are swapped out to
disk. They are faulted back into memory when accessed, resulting in significant
performance cost.  Charm++ uses a programming model where computations can be
scheduled by the language runtime. A designated thread can prefetch pages
required by a computation, averting page faults \cite{charmpp}. Rather than
implement page prefetching in a language runtime, one might implement a
prefetcher in the swapping subsystem, making it general purpose. A large body
of work already exists on hardware prefetching for processor caches, on which a
swap prefetcher might draw for inspiration \cite{prefetching}. However, this approach
can only work if page access patterns are predictable. Moreover, the high
latency of disks implies that prefetchers would have to predict the very
distant future (e.g. seconds ahead of execution). (TODO: Mike's comment on the
previous version of this paragraph said that it was ``too distant/off topic''.
Is it better now? I feel like prefetching was an idea that we considered too
much to leave out...)

@@

Finally, there have been studies which look at the performance overheads
associated with current implementations of virtual memory and suggest mechanisms
to mitigate them. RadixVM tries to overcome performance issues in highly
concurrent workloads due to serialization of memory management operations on
kernel data structures \cite{radixvm}. This work demonstrates that many parts of
existing memory management schemes are not scalable to larger systems.
Similarly, in our work, we wish to examine scalability limitations of memory
management in the Linux kernel. Other studies have suggested that \texttt{struct
page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 6 jokes in the paper?

>>>>

\begin{abstract}
    \small

large memory systems

study scalability of linux kernel as first step

point to some results

\end{abstract}

@@

Currently, there are few good tools available to systems researchers for
understanding how systems interact with large memories. An important first step
towards building such a tool is to identify bottlenecks in managing large
memories that make a tool difficult to develop.

In particular, we look at the case of a single process using a large amount of
memory in Linux. We run experiments to identify the scalability of the Linux
memory management system across multiple dimensions. First, we measure the
overhead of \texttt{vm\_area\_struct}, a common bookkeeping data structure in
the Linux kernel. Then, we measure the memory and processing overheads of
swapping. From these data, we extrapolate to systems running with more memory
with an eye on implications for studying large systems. While these experiments
are not comprehensive, we believe they can offer insights that are useful both
for building a research tool and for building systems.

@@

A process may use the \texttt{mmap} system call to request that the kernel
allocate part of its virtual address space. Accesses to memory addresses
outside of allocated regions result in a segmentation fault.
The Linux kernel keeps track of which portions of a process's address space
have been allocated using the \texttt{vm\_area\_struct} data structure. Because
processes may map a number of regions, including shared objects, stack, heap,
text, and anonymous regions, the scalability of this structure is critical to
Linux's performance.

@@

In our first set of experiments, we measure the memory overhead and latency of
the Linux Memory Management system's use of \texttt{vm\_area\_struct}, which
represents a region of a process's virtual address space. All the physical
pages in the system are managed using the \texttt{struct page} descriptors in
the kernel.  We do not measure the overhead of \texttt{struct page} because it
is known that there is one struct for each physical page. We do not measure the
overhead of page tables because this can be calculated from the size of the
virtual memory space used by the process.  When a program makes a memory
management system call such as \texttt{mmap}, \texttt{mremap}, \texttt{munmap},
\texttt{mprotect}, \texttt{mlock}, or \texttt{madvise}, the kernel is creating,
changing, or removing \texttt{vm\_area\_struct}s; thus, measuring the
scalability of Linux memory management requires measuring the overhead of these
structs.

@@

Before swapping a page out to disk, the kernel must reserve a slot for the page
in the swap space, unmap it from the process's address space by removing its
page table entries and flushing the TLB, and record which swap slot corresponds
to the page.

@@

The PFRA runs primarily as a kernel thread called kswapd. kswapd runs at regular
intervals and sleeps in between. Also, it may be invoked if the system detects
that memory has gone below a threshold. kswapd will continue to run until enough
pages are freed to bring the free memory pool above some threashold. Then,
kswapd goes back to sleep.  One concern with systems that have a large memory
usage is that kswapd may expend significant CPU time scanning through hundreds
of thousands of pages.

@@

% TODO: should we mention slab allocator?
To measure memory usage, we use the procfs's reporting on current memory usage.
Where appropriate, we record the number of active
\texttt{vm\_area\_struct}s in the system. Measuring kernel memory usage and
associating memory usage with a particular process is difficult. Our methodology assumes
that our benchmarks are running in an otherwise idle system and should dominate
increases in memory usage.  We compute total memory usage by the rest of the
system as the total amount of memory less the amount of memory used by the
benchmark. We keep transparent huge pages on throughout our experiments, unless
otherwise specified.

@@

First, when a continuous \texttt{mmap} (with the same permissions) happens, the
kernel memory allocator does not need to create a new \texttt{vm\_area\_struct}.
Instead, it uses the existing \texttt{vm\_area\_struct} to account for the
newly allocated space. This obviates the need to use the slab allocator to create
a new struct.

@@

Linux's use of \texttt{vm\_area\_struct} tends to be scalable in the general
case. The kernel latency of creating a memory region is lowest when the new
memory is allocated next to an already allocated region. This latency is the
worst in the \texttt{frag} microbenchmark i.e., when an already mapped region
is to be expanded.  We recommend that applications that manage their own memory
and expand memory regions should avoid this pattern.

\begin{figure}[t]
    \includegraphics[width=\columnwidth]{figures/mmap_latency}
    \caption{Latency of memory operations (TODO: improve the scale to make log behavior more apparent)}
    \label{fig:mmap_latency}
\end{figure}

@@

The \texttt{rand} workload, on the other hand, writes to pages that are likely not
adjacent and likely not subpages of a single huge page. As a result, a huge page
is allocated with almost every write, and system memory is exhausted and begins
swapping very quickly (20,000 operations compared to 4 million with the \texttt{cont} benchmark). Memory made available by kswapd is quickly consumed by
further allocations. Hence, the memory used by the workload continues to stay
close to its peak value. As expected, the behavior is the same irrespective of
the size of the swap file.

@@

The \texttt{cont} benchmark has more surprising results. Notice that Figure
\ref{fig:swap_time_cont} has a logarithmic y-scale. The system only begins to
swap after about 4M operations, as before. Unlike the \texttt{rand} benchmark, we see a
number of operations that take hundreds of milliseconds. Figure
\ref{fig:swap_time_cont_cdf} shows that these only form a fraction of a percent
of all accesses done after swapping starts. However, Figure
\ref{fig:swap_time_cdf} shows that these few accesses form a majority of run
time; they impact performance significantly.

@@

Mark was telling Suhas a TCP joke the other day. He had to keep repeating it slower
and slower until Suhas got it.

@@

I would tell you a UDP joke, but you might not get it. But if you did get it, you
wouldn't acknowledge it.

@@

TODO: in progress... should be ready for final product...

@@

Initially, we had intended to build a tool for studying large memory systems.
We found that the first step in doing so is studying the scalability of current
memory management systems themselves. Thus, our related work section, while
still relevant to our overall goal, is more directed at the second step of
building the tool.

@@

Likewise, the Quartz emulator studies the behaviour of applications on
Non-volatile Memory (NVM) while actually running on top of DRAM. It uses
existing hardware capabilities to emulate the higher latency and lower bandwidth
of most NVM technologies. While Quartz does not address the fact that the actual
memory available to the applications on NVM might be orders of magnitude higher
than on DRAM, it demonstrates that new technologies may be emulated using
existing technology \cite{quartz}.

Two threads walk into a bar. Bartender Hasnain looks up and yells, ``Hey, I want
don’t any conditions race like time last!''

@@

Interestingly, many of non-volatile memory technologies are making their ways
into storage drives. One potential use of fast storage devices is for fast
swapping space. In such a system, the overhead of swapping may become a
bottleneck.

Prefetching pages from swap space can offer a way to mitigate the overhead of
memory overcommitment, possibly making large virtual machines a viable
mechanism for studying large memory systems. When there is significant memory
pressure, even pages which are likely to be accessed soon are swapped out to
disk. They are faulted back into memory when accessed, resulting in significant
performance cost.  Charm++ uses a programming model where computations can be
scheduled by the language runtime. A designated thread can prefetch pages
required by a computation, averting page faults \cite{charmpp}. Rather than
implement page prefetching in a language runtime, one might implement a
prefetcher in the swapping subsystem, making it general purpose. A large body
of work already exists on hardware prefetching for processor caches, on which a
swap prefetcher might draw for inspiration \cite{prefetching}. However, this
approach can only work if page access patterns are predictable. Moreover, the
high latency of disks currently implies that prefetchers would have to predict
the very distant future (e.g. seconds ahead of execution); however, with
non-volatile memory-based swap space, this idea may be more practical.

@@

Finally, there have been studies which look at the performance overheads
associated with current implementations of virtual memory and suggest mechanisms
to mitigate them. RadixVM tries to overcome performance issues in highly
concurrent workloads due to serialization of memory management operations on
kernel data structures \cite{radixvm}. This work demonstrates that many parts of
existing memory management schemes are not scalable to larger systems.
Similarly, in our work, we wish to examine scalability limitations of memory
management in the Linux kernel. Other studies have suggested that \texttt{struct
page}, \texttt{struct vm\_area\_struct}, and page tables tend to comprise a
large portion of memory management overhead \cite{simics, struct_page}. In conclusion, did you find all 7 jokes in the paper?
